Compositional Quantitative Resource Analysis of C Programs

### 2014-06-21 paper/main.tex
new example for the stack metric
more work on the informal explanation

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been lot of progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
%
In this paper, we present a novel cost analysis system for C programs
that is based on a solid semantic foundation. It works automatically,
efficiently and precisely for many examples, can be used interactively
for manual bound derivation if the automation fails, and produces
verifiable certificates for the bounds.  Our contributions are as
follows.
%
\begin{enumerate}
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement which is parametric in a
  cost metric.  The semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost
  model.
\item We describe an automatic static analysis technique for deriving
  bounds for the cost that is defined by the operational semantics.
  It is based on amortized resource analysis for functional programs
  [XXX] and derives bounds that are non-negative linear combinations
  of sizes $|[x,y]|$ of intervals between program variables $x,y$.
\item We show with a publically available prototype implementation and
  an experimental evaluation that our novel amortized resource
  analysis works efficiently for realistic programs and that the
  derived constant factors in the bounds are close to or identical
  with the optimal ones.
\item We define a quantitative Hoare logic for interactively deriving
  cost bounds for C programs.  The logic is implemented and proved
  sound in the Coq Proof Assistant with respect to CompCert C
  light.  We show that our automatic analysis readily produces
  derivations in the quantitative Hoare logic.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
%
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  The quantitative Hoare logic and the automatic amortized
analysis enable semi-automatic interactive resource-bound development.
An automatically derived bound can be used in the quantitative logic
and a hand derived bound can be used by the automatic analysis.  Both
the automatic amortized analysis and the quantitative logic are
naturally compositional and describe the resource behavior of code
fragments without referring to the source code.  As a result, we
naturally derive global bounds that are functions of the input
parameters of the program.  To the best of our knowledge, our system
is the only one that can deal with resource like memory that may
become available during the execution.

@@

Following the development steps of automatic amortized analysis for
functional programs [XXX], we deliberately restrict our self to linear
bounds in the automatic analysis (there a no restrictions for the
manually-derived bounds in the logic).  More specifically, bounds have
the form $\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer
variables or constants.  The reason is mainly clarity of presentation,
since an extension to multivariate resource polynomials [XXX] would
make the inference rules considerably more involved and should
better be described separately.  However, we developed the linear
inference system so that the extension to polynomial bounds shall
work smoothly.

@@

Our new quantitative Hoare logic for CompCert Clight is a
generalization of a program logic that we recently developed to
formally very stack-space bounds in Coq [PLDI].  In a nutshell, we
replace the boolean predicates (pre and post conditions) of Hoare
logic with numeric predicates that map a program state to a
non-negative number or infinity.  The intuitive meaning of a
quantitative Hoare triple $\htriple{P}{C}{Q}$ that $C$ consumes $P(S)$
resources when executed in starting state $S$, and that $Q(S')$
resources are left if $C$ terminates in state $S'$.  Infinity
corresponds to \emph{false} in classic Hoare logic.  Our quantitative
logic is proved sound with respect to a cost semantic for Clight that
accounts a constant cost for each evaluation step.  This constant cost
is parametric in a resource metric that can be defined by the user.
Our shallow embedding in Coq enables the derivation of both, bounds
that are parametric in the cost metric or bounds for one specific
metric.

@@

The idea that drives the design of our system is amortized analysis
[XXX].  Assume that a program $p$ executes on a stating state
$\state_0$ and consumes $n$ resource units of some quantity we are
interested in.  We denote that by writing $(p, \state_0)
\Downarrow_n$.  The basic idea of amortized analysis is to define a
potential function $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
program state such that such that $(p, \state) \Downarrow_n$.

Since programs can be composed we must enrich the evaluation judgement
$(p, \state) \Downarrow_{n}$ to talk about the state resulting from
the program execution.  We write $(p, \state) \Downarrow_n \state'$ to
say that $\state'$ is the program state after the execution of $p$ in
starting state $\sigma$.  We now use two potential functions, one that
applies before the execution, and one that applies after.  The two
functions must respect the relation $\Phi(\state) \ge n +
\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(p,
\state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$ must
provide enough \emph{potential} for, both, paying for resource cost of
the computation and paying for the potential $\Phi'(\state')$ on the
resulting state $\state'$. That way, if $(\state, p_1) \Downarrow_n
\state'$ and $(\state', p_2) \Downarrow_m \state''$, we get
$\Phi(\state) \ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m +
\Phi''(\state'')$.  This can be composed to $\Phi(\state) \ge (n + m)
+ \Phi''(\state'')$.  Note that the initial potential function $\Phi$
provides an upper bound on the resource consumption of the whole
program.  What we have observed is that, if we define $\htriple{\Phi}
{p}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{p_1}{\Phi'} \\ \htriple{\Phi'}{p_2}{\Phi''}}
{\htriple{\Phi} {p_1; p_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect systems
rules.  This rules enable to interactively reason about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

We have implemented and proved sound such a quantitative Hoare for
CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
embedding in Coq, every function that can be expressed in Coq can in
principle be used as a resource bound.  We can use the logic to prove
a concrete bound for a given resource metric, or a bound that is
parametric in set of resource metrics.  It is also possible to
incorporate boolean preconditions into the bounds to express that the
bound is only valid for a certain class of inputs.  To this end, we
allow the potential function $\Phi$ in the pre and post conditions to
take non-negative number or $\infty$ (infinity) as values.  Infinity,
plays the same role as $\bot$ in Hoare logic.  Boolean assertions can
be embedded into potential functions by mapping $\bot$ to $\infty$ and
$\top$ to $0$.

@@

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

>>>>

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{itemize}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been lot of progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
%
In this paper, we present a novel cost analysis system for C programs
that is based on a solid semantic foundation. It works automatically,
efficiently and precisely for many examples, can be used interactively
for manual bound derivation if the automation fails, and produces
verifiable certificates for the bounds.  Our contributions are as
follows.
%
\begin{enumerate}
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement.  The cost is parametric
  in a cost metric and the semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost model.

\item We describe an automatic static analysis technique for deriving
  bounds for the cost that is defined by the operational semantics.
  It is based on amortized resource analysis for functional
  programs~\cite{Jost03,HoffmannAH12} and derives bounds that are
  non-negative linear combinations of sizes $|[x,y]|$ of intervals
  between program variables $x,y$.
\item We show with a publically available prototype implementation and
  an experimental evaluation that our novel amortized resource
  analysis works efficiently for realistic programs and that the
  derived constant factors in the bounds are close to or identical
  with the optimal ones.
\item We define a quantitative Hoare logic for interactively deriving
  cost bounds for C programs.  The logic is implemented and proved
  sound in the Coq Proof Assistant with respect to CompCert C
  light.  We show that our automatic analysis readily produces
  derivations in the quantitative Hoare logic.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
%
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  The quantitative Hoare logic and the automatic amortized
analysis enable semi-automatic interactive resource-bound development.
An automatically derived bound can be used in the quantitative logic
and a hand derived bound can be used by the automatic analysis.  Both
the automatic amortized analysis and the quantitative logic are
naturally compositional and describe the resource behavior of code
fragments without referring to the source code.  As a result, we
naturally derive global bounds that are functions of the input
parameters of the program.  To the best of our knowledge, our system
is the only one that can deal with resource like memory that may
become available during the execution.

@@

Following the development steps of automatic amortized analysis for
functional programs~\cite{Jost03,HoffmannH10}, we deliberately
restrict our self to linear bounds in the automatic analysis (there a
no restrictions for the manually-derived bounds in the logic).  More
specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
where $a$ and $b$ are integer variables or constants.  The reason is
mainly clarity of presentation, since an extension to multivariate
resource polynomials~\cite{HoffmannAH11} would make the inference rules considerably
more involved and should better be described separately.  However, we
developed the linear inference system so that the extension to
polynomial bounds shall work smoothly.

@@

Our new quantitative Hoare logic for CompCert Clight is a
generalization of a program logic that we developed to formally very
stack-space bounds for C programs in Coq~\cite{veristack14}.  In a
nutshell, we replace the boolean predicates (pre and post conditions)
of Hoare logic with numeric predicates that map a program state to a
non-negative number or infinity.  The intuitive meaning of a
quantitative Hoare triple $\htriple{P}{S}{Q}$ that $C$ consumes $P(\state)$
resources when executed in starting state $\state$, and that $Q(\state')$
resources are left if $C$ terminates in state $\state'$.  Infinity
corresponds to \emph{false} in classic Hoare logic.  Our quantitative
logic is proved sound with respect to a cost semantic for Clight that
accounts a constant cost for each evaluation step.  This constant cost
is parametric in a resource metric that can be defined by the user.
Our shallow embedding in Coq enables the derivation of both, bounds
that are parametric in the cost metric or bounds for one specific
metric.

@@

The idea that drives the design of our system is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$
executes on a starting state $\state_0$ and consumes $n$ resource units
of some quantity we are interested in.  We denote that by writing $(S,
\state_0) \Downarrow_n$.  The basic idea of amortized analysis is to
define a potential function $\Phi$ that maps program states to
non-negative numbers and to show that $\Phi(\state) \geq n$ if
$\state$ is a program state such that such that $(S, \state)
\Downarrow_n$.

Since programs can be composed we must enrich the evaluation judgement
$(S, \state) \Downarrow_{n}$ to talk about the state resulting from
the program execution.  We write $(S, \state) \Downarrow_n \state'$ to
say that $\state'$ is the program state after the execution of $S$ in
starting state $\sigma$.  We now use two potential functions, one that
applies before the execution, and one that applies after.  The two
functions must respect the relation $\Phi(\state) \ge n +
\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(S,
\state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$ must
provide enough \emph{potential} for both, paying for resource cost of
the computation and paying for the potential $\Phi'(\state')$ on the
resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get
$\Phi(\state) \ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m +
\Phi''(\state'')$.  This can be composed to $\Phi(\state) \ge (n + m)
+ \Phi''(\state'')$.  Note that the initial potential function $\Phi$
provides an upper bound on the resource consumption of the whole
program.  What we have observed is that, if we define $\htriple{\Phi}
{S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect systems
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

We have implemented and proved sound such a quantitative Hoare for
CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
embedding in Coq, every function that can be expressed in Coq can in
principle be used as a resource bound.  We can use the logic to prove
a concrete bound for a given resource metric, or a bound that is
parametric in a set of resource metrics.  It is also possible to
incorporate boolean preconditions into the bounds to express that the
bound is only valid for a certain class of inputs.  To this end, we
allow the potential function $\Phi$ in the pre and post conditions to
take non-negative number or $\infty$ (infinity) as values.  Infinity,
plays the same role as $\bot$ in Hoare logic.  Boolean assertions can
be embedded into potential functions by mapping $\bot$ to $\infty$ and
$\top$ to $0$.

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

@@
